### 2024-11-13 Article/Background_sections/Adhesivity.tex
minori

\begin{definition}[$\mathcal{M}$-adhesivity]\label{def:adh}
    Let $\cat C$ be a category and $\mathcal M \subseteq \Mono(\cat C)$ containing all isomorphisms, closed under composition and stable under pullbacks and pushouts (\Cref{def:stab_under_pb_po}).
    Then $\cat C$ is \emph{$\mathcal M$-adhesive} if
    \begin{enumerate}
        \item every cospan $C \xrightarrow[]{g} D \xleftarrow[]{m} B$ with $m \in \mathcal M$ can be completed to a pullback (such pullbacks are called $\mathcal M$-pullbacks);
        \item every span $C \xleftarrow{m} A \xrightarrow{f} B$ with $ m \in \mathcal M$ can be completed to a pushout (such pushouts are called $\mathcal M$-pushouts);
        \item pushouts along $\mathcal M$-arrows are $\mathcal M$-Van Kampen squares.
    \end{enumerate}
    We also say that $\cat C$ is \emph{adhesive} when it is $\Mono(\cat C)$-adhesive, and \emph{quasiadhesive} when it is $\Reg(\cat C)$-adhesive.
\end{definition}

@@

\begin{prop}\label{prop:monos_are_preserved_by_pullbacks_in_adh_cats}
    In any adhesive category, the pushout of a monomorphism along any morphism is a monomorphism, and the resulting square is also a pullback.
\end{prop}

@@

\begin{cor}\label{cor:adhesivity_functor_categories}
    Let $\cat A$ be a $\mathcal M$-adhesive category for some $\mathcal M \subseteq \Hom(\cat A)$. Then, for every other category $\cat C$, the functor category $[\cat C, \cat A]$ is $\mathcal M^{\cat C}$-adhesive, where
    \[
        \mathcal{M}^\cat C = \{\eta \in \Hom([\cat C, \cat A]) \mid \eta_C \in \mathcal M \textit{ for each object $C$ of $\cat C$ }\}
    \]
\end{cor}

@@

\begin{lemma}\label{lemma:pushouts_kernel_pairs}
    Let $\cat C$ be an $\mathcal{M}$-adhesive category with all pullbacks, and suppose that the cube below is given, in which every face is a pullback and the bottom one is a $\mathcal{M}$-pushout.
        \[\begin{tikzcd}[row sep=25, column sep=25]
	& {A'} && {B'} \\
	{C'} && {D'} \\
	& A && B \\
	C && D
	\arrow["{f'}", from=1-2, to=1-4]
	\arrow["{m'}"', from=1-2, to=2-1]
	\arrow["a"'{pos=0.7}, from=1-2, to=3-2]
	\arrow["{n'}"', from=1-4, to=2-3]
	\arrow["b", from=1-4, to=3-4]
	\arrow["{g'}"{pos=0.7}, from=2-1, to=2-3, crossing over]
	\arrow["c"', from=2-1, to=4-1]
	\arrow["f"'{pos=0.3}, from=3-2, to=3-4]
	\arrow["d"{pos=0.3}, from=2-3, to=4-3, crossing over]
	\arrow["m", from=3-2, to=4-1]
	\arrow["n", from=3-4, to=4-3]
	\arrow["g"', from=4-1, to=4-3]
    \end{tikzcd}\]

>>>>

\begin{definition}[$\mathcal{M}$-adhesivity]\label{def:adh}
    Let $\cat C$ be a category and $\mathcal M \subseteq \Mono(\cat C)$ containing all isomorphisms, closed under composition and stable under pullbacks and pushouts (\Cref{def:stab_under_pb_po}).
    Then $\cat C$ is \emph{$\mathcal M$-adhesive} if
    \begin{enumerate}
        \item every cospan $C \xrightarrow[]{g} D \xleftarrow[]{m} B$ with $m \in \mathcal M$ can be completed to a pullback (such pullbacks are called $\mathcal M$-pullbacks);
        \item every span $C \xleftarrow{m} A \xrightarrow{f} B$ with $ m \in \mathcal M$ can be completed to a pushout (such pushouts are called $\mathcal M$-pushouts);
        \item pushouts along $\mathcal M$-arrows are $\mathcal M$-Van Kampen squares.
    \end{enumerate}
    We also say that $\cat C$ is \emph{adhesive} when it is $\Mono(\cat C)$-adhesive, and \emph{quasiadhesive} when it is $\Reg(\cat C)$-adhesive. \todo{NO, questo è sbagliato: bisogna introdurre la nozione di adesività forte (cioè VK vale per tutti i cubi)}
\end{definition}

@@

\begin{prop}\label{prop:monos_are_preserved_by_pullbacks_in_adh_cats}
    In any adhesive category, the pushout of a monomorphism along any morphism is a monomorphism, and the resulting square is also a pullback. \todo{almeno la seconda parte devi provarla o citarla per le M-adesive}
\end{prop}

@@

\begin{cor}\label{cor:adhesivity_functor_categories}
    Let $\cat A$ be a $\mathcal M$-adhesive category for some $\mathcal M \subseteq \Hom(\cat A)$. Then, for every other category $\cat C$, the functor category $[\cat C, \cat A]$ is $\mathcal M^{\cat C}$-adhesive, where \todo{Devi chiedere che C abbia tutti i pullback}
    \[
        \mathcal{M}^\cat C = \{\eta \in \Hom([\cat C, \cat A]) \mid \eta_C \in \mathcal M \textit{ for each object $C$ of $\cat C$ }\}
    \]
\end{cor}

@@

\color{blue}\todo{Ti inserisco una versione un po' più generale del lemma sotto}

\begin{lemma}\label{lemma:pushouts_kernel_pairs}
	Let $\cat C$ be an \todo{Qua serve adesività forte} $\mathcal{M}$-adhesive category with all pullbacks, and suppose that in the cube below the top face is an $\mathcal{M}$-pushout.
	\[\begin{tikzcd}[row sep=25, column sep=25]
		& {A'} && {B'} \\
		{C'} && {D'} \\
		& A && B \\
		C && D
		\arrow["{f'}", from=1-2, to=1-4]
		\arrow["{m'}"', from=1-2, to=2-1]
		\arrow["a"'{pos=0.7}, from=1-2, to=3-2]
		\arrow["{n'}"', from=1-4, to=2-3]
		\arrow["b", from=1-4, to=3-4]
		\arrow["{g'}"{pos=0.7}, from=2-1, to=2-3, crossing over]
		\arrow["c"', from=2-1, to=4-1]
		\arrow["f"'{pos=0.3}, from=3-2, to=3-4]
		\arrow["d"{pos=0.3}, from=2-3, to=4-3, crossing over]
		\arrow["m", from=3-2, to=4-1]
		\arrow["n", from=3-4, to=4-3]
		\arrow["g"', from=4-1, to=4-3]
	\end{tikzcd}\]

\begin{proof}
	Since $f'$, being the pullback of $f$, is in $\mathcal{M}$ then by  \Cref{prop:monos_are_preserved_by_pullbacks_in_adh_cats} the bottom face of the cube is a pullback. Thus \Cref{lemma:kern_pairs_pres_pullbacks} entails that in the following cube the vertical faces are pullbacks \todo{Nota che le frecce verticali NON sono mono in generale: qua serve adesività forte}.
	\[\begin{tikzcd}[row sep = 20, column sep = 20]
		& {K_a} && {K_b} \\
		{K_c} && {K_d} \\
		& {A'} && {B'} \\
		{C'} && {D'}
		\arrow["{k_{f'}}", from=1-2, to=1-4]
		\arrow["{k_{m'}}"', from=1-2, to=2-1]
		\arrow["{\pi_a^1}"{pos=0.7}, from=1-2, to=3-2]
		\arrow["{k_{n'}}", from=1-4, to=2-3]
		\arrow["{\pi_b^1}", from=1-4, to=3-4]
		\arrow["{k_{g'}}"{pos=0.7}, from=2-1, to=2-3, crossing over]
		\arrow["{\pi_c^1}"', from=2-1, to=4-1]
		\arrow["{f'}"'{pos=0.3}, from=3-2, to=3-4]
		\arrow["{\pi_d^1}"{pos=0.7}, from=2-3, to=4-3, crossing over]
		\arrow["{m'}", from=3-2, to=4-1]
		\arrow["{n'}", from=3-4, to=4-3]
		\arrow["{g'}"', from=4-1, to=4-3]
	\end{tikzcd}\]

Now the thesis follows from strong $\mathcal{M}$-adhesivity.
\end{proof}

\color{black}

\begin{lemma}\label{lemma:pushouts_kernel_pairs}
    Let $\cat C$ be an \todo{Qua serve adesività forte} $\mathcal{M}$-adhesive category with all pullbacks, and suppose that the cube below is given, in which every face is a pullback and the bottom one is a $\mathcal{M}$-pushout.
        \[\begin{tikzcd}[row sep=25, column sep=25]
	& {A'} && {B'} \\
	{C'} && {D'} \\
	& A && B \\
	C && D
	\arrow["{f'}", from=1-2, to=1-4]
	\arrow["{m'}"', from=1-2, to=2-1]
	\arrow["a"'{pos=0.7}, from=1-2, to=3-2]
	\arrow["{n'}"', from=1-4, to=2-3]
	\arrow["b", from=1-4, to=3-4]
	\arrow["{g'}"{pos=0.7}, from=2-1, to=2-3, crossing over]
	\arrow["c"', from=2-1, to=4-1]
	\arrow["f"'{pos=0.3}, from=3-2, to=3-4]
	\arrow["d"{pos=0.3}, from=2-3, to=4-3, crossing over]
	\arrow["m", from=3-2, to=4-1]
	\arrow["n", from=3-4, to=4-3]
	\arrow["g"', from=4-1, to=4-3]
    \end{tikzcd}\]
